Methods that remove batch effects while retaining group differences may lead to exaggerated confidence in downstream analyses

### 2014-11-21 tex/manuscript.tex
Oppdatert Figure 1 samt diskusjon
Diff mot versjon fra 17. november.

Problems with batch adjusting unbalanced data sets are not specific to ComBat, but are shared by other batch adjustment methods: indeed, the empirical Bayes approach used by ComBat  will dampen the batch adjustments and may thus reduce the problem slightly relative to a more direct two-way ANOVA approach. Although some of the tools warn against performing subsequent analyses on batch adjusted data, these warnings are often somewhat vague and seem to be primarily concerned about the degrees of freedom in the data set. Batch adjustment will reduce the degrees of freedom in the data which may influence the distribution of the test statistics in subsequent analyses, e.g. $F$-statistics from ANOVA. For small data sets or where there are many small batches this may be a problem, while it would be less of a concern when there are a fair number of samples in each batch. However, we find that the main problem is not related to the degrees of freedom, but estimation errors during the batch adjustments which get applied across the data, which can induce group differences in the adjusted data set even in the absence of batch effects.

@@

For an investigator facing an unbalanced data set with batch effects, our primary advice would be to adjust for batch inside the statistical test, and avoid the two-step procedure outlined above. If this is not possible, batch correction using outcome as covariate should only be performed with great caution, and downstream confidence estimates must be treated with suspicion.

We did consider if batch adjustment could be modified so as to balance the need for removing batch effects with avoidance of introducing false group differences which would be somewhere between batch centering and two-way ANOVA based adjustment. For two-group comparisons, the batch centering will remove some of the group effects and thus underestimate group differences, while two-way ANOVA based batch adjustment will add random, but unbiased, group effects. Somewhere in between is a balance for which these two balance out. However, the applicability of this approach is very limited and potentially risky since the degree of moderation of the batch adjustment will depend on the subsequent comparison to be made.

@@

Knowing that adjustment for batch effects while preserving the group difference may lead to varying degree of false results, to what extent can an investigator trust published results where such a method has been applied? Essentially, when the batch--group configuration is balanced, or group difference is ignored  (i.e. group labels not given as parameters to ComBat), problems related to preserving group differences will not occur. For other cases, a re-analysis without using this approach is the most rigorous path. However, such a re-analysis is not feasible if the downstream  analysis can not independently adjust for batch effects. To reach a reliable result, batch effects need to be handled in some way or another. To make matters worse, a re-analysis relies on the availability of the raw data and a description of processing and analysis steps taken in the original work. Even when this is available, the necessary statistics and bioinformatics skills and work hours could still be in short supply.

A superficial assessment of the reliability of results based on batch corrected data with group differences retained can often be made. A major concern would be batches where either of the groups of interest are missing or strongly under-represented, which would contribute to the nominal sample size, but not to the effective sample size. In section~\ref{sec:twogroups}, the ratio $\nu_0/\nu\ge1$ indicates by how much the effective sample size is overestimated when comparing two groups. A similar computation, although more complicated, can be made for multiple groups: see appendix for details covering general linear models. However, the two-group assessment should still give a fair idea of the reliability of pairwise group comparisons. If the $\nu_0/\nu$ ratio is close to 1, results should still be reliable, while a ratio much larger than 1 will indicate exaggerated confidence. If the behaviour of the test statistic is well understood, as the $F$-statistic in ANOVA, one may adjust the statistic according to the $\nu_0/\nu$ ratio and recompute the $p$-value. However, this will require deep insight into the statistical behaviour of the model. Down-sampling to the effective samples size, i.e. to a $\nu/\nu_0$ portion of the samples, may also be done to see if results persist. However, none of these solutions are ideal.

@@

\begin{figure}[!p]
\centering\includegraphics[width=8cm]{Fig/boxplots.pdf}
%\centering\includegraphics[width=13cm]{Fig/boxplots_old.pdf}% For the old version
%\centering\includegraphics[width=13cm]{Fig/boxplots.png}
\caption{We simulated expression of one gene from four study groups unevenly distributed in three batches containing 50+0+0+20, 0+50+0+20 and 0+0+50+20 samples from groups 1 to 4. This case, design and effects, was selected to illustrate the spurious effects that may arise from different batch adjustments. The Y-axis represent the expression values, while the X-axis is used to visually separate the batches. Circles, triangles, and crosses indicate values from each of the three batches, with colours indicating study groups. Correspondingly coloured boxes to the right of the measurements show group means with 95\% confidence intervals. These give some indication as to which group differences would be found significant.
a) The ``true'' values measured in a system without batch effects. Groups 1 and 2 have lower means than groups 3 and 4.
b) The same samples as in a) but measured in a system with batch effects. All the groups seem different.
c) The measurements in b) are adjusted with batch centering. Group 3 and 4 seem to differ while group 1 and 3 are more similar.
d) The measurements in b) are adjusted with two-way ANOVA based batch centering (using \texttt{limma}). Group 1 and 2 seem to differ.
e) The least squares estimates of the group means from a two-way ANOVA have the same means as in d), but more appropriate confidence intervals.
}
\label{fig:boxplots}
\end{figure}

>>>>

Problems with batch adjusting unbalanced data sets are not specific to ComBat, but are shared by other batch adjustment methods: indeed, the empirical Bayes approach used by ComBat  may dampen the batch adjustments and thereby reduce the problem slightly relative to a more direct two-way ANOVA approach. Although some of the tools warn against performing subsequent analyses on batch adjusted data, these warnings are often somewhat vague and seem to be primarily concerned about the degrees of freedom in the data set. Batch adjustment will reduce the degrees of freedom in the data which may influence the distribution of the test statistics in subsequent analyses, e.g. the variance of the $F$-statistics from ANOVA. For small data sets or where there are many small batches this may be a problem, while it would be less of a concern when there are a fair number of samples in each batch. However, we find that the main problem is a systematic bias, e.g. scaling the $F$-statistic by a factor, induced as estimation errors during the batch adjustments are applied across the data, which can induce group differences in the adjusted data set even in the absence of batch effects.

@@

For an investigator facing an unbalanced data set with batch effects, our primary advice would be to adjust for batch as part of the statistical analysis, and avoid the two-step procedure outlined above. If this is not possible, batch correction using outcome as covariate should only be performed with great caution, and downstream confidence estimates must be treated with suspicion. Comparing results from batch adjusted data with and without covariates may indicate to what extent the outcome depends on the mode of batch adjustment.

We did consider if a partial batch adjustment could be made to balance the need for removing batch effects against the risk of introducing false group differences which would be somewhere between batch centering and two-way ANOVA based adjustment. For two-group comparisons, the batch centering will remove some of the group effects and thus underestimate group differences, while two-way ANOVA based batch adjustment will add random, but unbiased, group effects. Somewhere in between is a balance for which the effect of these two balance out in the statistical tests. However, the applicability of this approach is very limited and potentially risky since the appropriate degree of batch adjustment will depend on the subsequent comparison to be made.

@@

Knowing that adjustment for batch effects while preserving the group difference may lead to varying degree of false results, to what extent can an investigator trust published results where such a method has been applied? Essentially, when the batch--group configuration is balanced, batch adjustment will not bias the group difference estimates. For unbalanced designs, however, batch effects may act as a confounder, and a re-analysis without relying on the two-step approach is the most rigorous path. However, such a re-analysis may be infeasible or impractical, depending on the availability of the raw data and proper method description, the ability of analysis tools to handle batch effects. To reach a reliable result, batch effects need to be handled in some way or another, and the necessary statistics and bioinformatics resources.

A superficial assessment of the reliability of results based on batch corrected data with group differences retained can often be made. A major concern would be batches where either of the groups of interest are missing or strongly under-represented, which would contribute to the nominal sample size, but not to the effective sample size. In section~\ref{sec:twogroups}, the ratio $\nu_0/\nu\ge1$ indicates by how much the effective sample size is overestimated when comparing two groups. A similar computation, although more complicated, can be made for multiple groups: see appendix for details covering general linear models. If the $\nu_0/\nu$ ratio is close to 1, results should still be reliable, while a ratio much larger than 1 will indicate exaggerated confidence. If the behaviour of the test statistic is well understood, as the $F$-statistic in ANOVA, one may adjust the statistic according to the $\nu_0/\nu$ ratio and recompute the $p$-value. However, this will require deep insight into the statistical behaviour of the model. Down-sampling to the effective samples size, i.e. to a $\nu/\nu_0$ portion of the samples, may also be done to see if results persist. However, none of these solutions are ideal.

@@

\begin{figure}[!p]
\centering\includegraphics[width=14.5cm]{Fig/boxplots.pdf}
\caption{We simulated expression of one gene from three study groups unevenly distributed in three batches, designed to illustrate the spurious effects that may arise from different batch adjustments. The Y-axis represent the expression values, while the X-axis is used to visually separate the groups and batches. Circles, triangles, and crosses indicate values from each of the three groups. Surrounding boxes indicate batches. For each group, the coloured bar indicates mean and its 95\%~confidence interval, which give some indication as to which group differences would be found significant.
a) The ``true values'' measured in a system without batch effects. The first two groups are drawn from normal distributions with the same mean and variance, while the third has a lower mean.
b) Batch effects have been added to the ``true values'', as indicated by the shifting of the batch frames, leading to apparent differences between the first two groups.
c) The measurements in b) are adjusted by mean-centering each batch. All groups appear significantly different.
d) The measurements in b) are adjusted with two-way ANOVA based batch centering (using \texttt{limma}). Again, all groups appear to be significantly different.
e) The least squares estimates of the group means from a two-way ANOVA have the same means as in d), but more appropriate confidence intervals.
}
\label{fig:boxplots}
\end{figure}
